greta.Rmd

### 2019-09-23 manuscript/manuscript.Rmd
WIP
Need to revise & consolidate notebooks

---
title: "Long transients can mislead model inference and decision making -- and what to do about it"
date: "`r Sys.Date()`"
author:
  - name: Carl Boettiger
    email: cboettig@berkeley.edu
    affiliation: ucb
    footnote: Corresponding Author
address:
  - code: ucb
    address: "ESPM Department, University of California, 130 Mulford Hall Berkeley, CA 94720-3114, USA"
abstract: |
TBD
journal: TBD
layout: 5p
bibliography: refs.bib
output: rticles::elsevier_article

@@

```{r include = FALSE}
knitr::opts_chunk$set(echo = FALSE, cache = 1,  dev = "cairo_pdf", fig.width=4.5, fig.height=3.5)
library(hrbrthemes)
library(ggplot2)
library(Cairo)
library(extrafont)
extrafont::loadfonts()
```

@@

Periods of long transient dynamics have long been identified in the ecological systems.   Ecological @Hastings2018

>>>>

---
title: "Consequences of long transients in model inference and optimal control"
date: "`r Sys.Date()`"
author:
  - name: Carl Boettiger
    email: cboettig@berkeley.edu
    affiliation: ucb
    footnote: Corresponding Author
address:
  - code: ucb
    address: "ESPM Department, University of California, 130 Mulford Hall Berkeley, CA 94720-3114, USA"
abstract: |
  I explore the impact of long transient dynamics for model inference and optimal
journal: Theoretical ecology
layout: 5p
bibliography: refs.bib
output: rticles::elsevier_article

@@

```{r include = FALSE}
knitr::opts_chunk$set(echo = FALSE, cache = 1,  
                      dev = "cairo_pdf", fig.width=4.5, 
                      fig.height=3.5, message = FALSE,
                      warning = FALSE)
library(hrbrthemes)
library(ggplot2)
library(ggthemes)
library(Cairo)
library(extrafont)
extrafont::loadfonts()
```

@@

```{r}
source("../R/ghost.R")
```

```{r}
data <- vroom::vroom("../data/reps.csv.xz") %>% filter(t < 3000)
## get tail cases for comparison
ex <- data %>% group_by(reps) %>% summarise(ave = mean(x)) %>% arrange(ave)
lows <- ex %>% dplyr::slice(1:10) %>% pull(reps)
highs <- ex %>% dplyr::slice((n()-9):n()) %>% pull(reps)
```

```{r arima_forecast}
library(stats)
x <- data %>% filter(reps == lows[[6]], t < 2000) %>% pull(x) %>% as.ts() 
fit <- arima(x)
y <- predict(fit, n.ahead = 1000) %>% as.data.frame()
y <- bind_rows(data.frame(pred = x, se = 0), y)
arima_forecast <- data %>% filter(reps == lows[[6]]) %>% select(t,x) %>% bind_cols(y)
```

```{r arima_figure, message=FALSE, warning=FALSE}
train <- arima_forecast %>% filter(t < 2000) 
test <- arima_forecast %>% filter(t > 2000) 
train %>%
  ggplot(aes(t)) + 
  geom_line(aes(y=x)) +
  geom_line(aes(y=pred), lwd = 1, lty = 3, col = "red", data = test) + 
  geom_ribbon(aes(y=pred, ymin=pred-se, ymax=pred+se), alpha = 0.4, fill = "red", data = test)  + 
  geom_line(aes(y=x), data = test, lty = 2, col="blue") +
  xlim(0,3000) + ylim(0, 1.6)+ ylab("x") + theme_minimal()

```

Transient dynamics have long been identified in the ecological systems @Hastings1994.   Ecological @Hastings2018

### 2019-09-23 notebook/ghost-fit-forecast.Rmd
WIP
Need to revise & consolidate notebooks

```{r}
data <- read_csv("../data/reps.csv.xz") %>% filter(t < 2000)
data %>% ggplot(aes(t,x, group=reps)) + geom_line(alpha=0.02)
```

Divide this data into several groups based on

```{r}
means <- data %>% group_by(reps) %>% summarise(ave = mean(x)) %>% arrange(ave)

lows <- means %>% dplyr::slice(1:10) %>% pull(reps)
highs <- means %>% dplyr::slice((n()-9):n()) %>% pull(reps)
write_csv(data.frame(lows=lows,highs=highs), "../data/extremes.csv.xz")

coding <- function(reps) case_when(
  reps %in% lows ~ "low",
  reps %in% highs ~ "high"
)
data %>% 
  mutate(coding = coding(reps)) %>% 
  na.omit() %>% 
ggplot(aes(t, x, group=reps, color = coding)) + geom_line(alpha=0.5)

#means %>% ggplot(aes(ave)) + geom_histogram(binwidth = 0.02)

@@

## highs

```{r}
wide <- data %>% 
  filter(reps %in% highs) %>%
  spread(reps, x) %>%
  select(-t) %>% 
  as.matrix()
n <- dim(wide)[1]

```{r}
x_t1 <- wide[-1,]
x_t <- wide[-n,] 
```

@@

write_csv(samples, "../data/posterior_highs.csv.xz")

@@

----

## lows

```{r}
wide <- data %>% 
  filter(reps %in% lows) %>%
  spread(reps, x) %>%
  select(-t) %>% 
  as.matrix()
n <- dim(wide)[1]

```{r}
x_t1 <- wide[-1,]
x_t <- wide[-n,] 
```

```{r}
a <- uniform(0, 1)
mean <- x_t + r * x_t * (1 - x_t / K) - a * x_t ^ Q / (x_t ^ Q + H ^ Q)
distribution(x_t1) <- normal(mean, sigma * x_t)
m <- model(a)
```

```{r }
system.time({
  draws2 <- mcmc(m, n_samples = 1000, warmup = 3000, chains = 4, verbose = FALSE)
})
```

```{r}
bayesplot::mcmc_trace(draws2)
```

```{r}
samples2 <-  
  map_dfr(draws2, 
          function(x) data.frame(x, t = 1:dim(x)[1]), 
          .id = "chain") %>% 
  gather(variable, value, -t, -chain)

samples2 %>% ggplot() + 
  geom_histogram(aes(value), bins = 30)  +
  geom_vline(data = true, aes(xintercept = value), col = "red", lwd = 1) + 
  facet_wrap(~variable, scales = "free")

write_csv(samples2, "../data/posterior_lows.csv.xz")
```

## Full ensemble

```{r}
wide <- data %>% 
  spread(reps, x) %>%
  select(-t) %>% 
  as.matrix()
n <- dim(wide)[1]

```{r}
x_t1 <- wide[-1,]
x_t <- wide[-n,] 
```

```{r}
a <- uniform(0, 1)
mean <- x_t + r * x_t * (1 - x_t / K) - a * x_t ^ Q / (x_t ^ Q + H ^ Q)
distribution(x_t1) <- normal(mean, sigma * x_t)
m <- model(a)
```

```{r }
system.time({
  draws3 <- mcmc(m, n_samples = 1000, warmup = 3000, chains = 6, verbose = FALSE)
})
```

```{r}
bayesplot::mcmc_trace(draws3)
```

```{r}
samples3 <-  
  map_dfr(draws3, 
          function(x) data.frame(x, t = 1:dim(x)[1]), 
          .id = "chain") %>% 
  gather(variable, value, -t, -chain)

samples3 %>% ggplot() + 
  geom_histogram(aes(value), bins = 30)  +
  geom_vline(data = true, aes(xintercept = value), col = "red", lwd = 1) + 
  facet_wrap(~variable, scales = "free")

write_csv(samples3, "../data/posterior_ensemble.csv.xz")
```

```{r }
p <- list(r = .05, K = 2, Q = 5, H = .38, sigma = .02, a=0.023, N = 4e3, x0 = 0.2, N = 1e4)
source("../R/ghost.R")
p$a <- 0.021
tf <- theory(p) %>% mutate(a = 0.021)

p$a <- 0.023
tf <- theory(p) %>% mutate(a = 0.023) %>% bind_rows(tf)

p$a <- 0.024
tf <- theory(p) %>% mutate(a = 0.024) %>% bind_rows(tf)

tf <- tf %>% mutate(a = as.factor(a))

tf %>%
  ggplot(aes(x, potential, col=a)) +
  geom_line(lwd=1)

```{r}
```

>>>>

```{r}
source("../R/ghost.R")
```

```{r}
data <- vroom::vroom("../data/reps.csv.xz") %>% filter(t < 3000)
## get tail cases for comparison
ex <- data %>% group_by(reps) %>% summarise(ave = mean(x)) %>% arrange(ave)
lows <- ex %>% dplyr::slice(1:10) %>% pull(reps)
highs <- ex %>% dplyr::slice((n()-9):n()) %>% pull(reps)
```

```{r arima_forecast}
library(stats)
x <- data %>% filter(reps == lows[[6]], t < 2000) %>% pull(x) %>% as.ts() 
fit <- arima(x)
y <- predict(fit, n.ahead = 1000) %>% as.data.frame()
y <- bind_rows(data.frame(pred = x, se = 0), y)
arima_forecast <- data %>% filter(reps == lows[[6]]) %>% select(t,x) %>% bind_cols(y)
```

```{r arima_figure, message=FALSE, warning=FALSE}
train <- arima_forecast %>% filter(t < 2000) 
test <- arima_forecast %>% filter(t > 2000) 
train %>%
  ggplot(aes(t)) + 
  geom_line(aes(y=x)) +
  geom_line(aes(y=pred), lwd = 1, lty = 3, col = "red", data = test) + 
  geom_ribbon(aes(y=pred, ymin=pred-se, ymax=pred+se), alpha = 0.4, fill = "red", data = test)  + 
  geom_line(aes(y=x), data = test, lty = 2, col="blue") +
  xlim(0,3000) + ylim(0, 1.6)+ ylab("x") + theme_minimal()

@@

```{r}
wide <- select(train, x) %>% as.matrix() 
n <- dim(wide)[1]
x_t1 <- wide[-1,]
x_t <- wide[-n,] 
```

@@

@@
